Using language as an indicator of cognitive decline in pre-clinical Alzheimer's Disease.

### 2020-02-20 Source/Paper 1 - Presidents Corpus Paper/PresidentsCorpus.tex
updated Presidentspaper

% Stronger motivation

@@

@@

\subsection{Longitudinal Analysis}
As with Berisha and Liss\cite{Berisha2015}, we analysed the transcripts over time in order to determine if there was an underlying temporal trend that may indicate that different parts of language increase or decrease over time. In AD, cognitive abilities are said to deteriorate over time and therefore it may be useful to analyse temporal trends that indicate a potential deterioration of cognition. In order to get a accurate measure of deterioration over time, we time stamped each transcript in relation to number of days from the first transcript to the date of the transcript being analysed as the press conferences were not evenly spaced out during the presidencies.
\par
We looked also compared the results of linear models and a non linear model in terms of best fit. The non linear model we used was the generalised additive model \cite{Hastie1986}.
There is evidence to support the idea that language deterioration. In figure 1, we look at the difference of between a linear model of Reagan's use of personal pronouns vs a non linear model. Whilst they are broadly similar in that they track an increase in his use of personal pronouns, they do not capture the variation in this increase as closely as a non linear model might. We therefore decided to apply this to all the features we extracted.

@@

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{3cm} | p{2.2cm} | p{2.2cm} | p{2.2cm} |}
		\hline
		& RR v GHWB & RR v DJT & GHWB v DJT \\ \hline
		Total Words & 6.6751 (574.8808, 1057.5324) & 5.352 (974.6629, 2175.8489) & 2.4774 (137.5811, 1380.5176) \\ \hline
		Unique Stems & 10.878 (161.1580, 232.6168) & 10.11 (275.4688, 414.0975) & 4.148 (75.86722, 219.92432) \\ \hline
		Mean Length of Utterance & 16.175 (3.915869, 5.009114) & 25.084 (8.453404, 9.922328)  & 13.484 (4.021601, 5.429148) \\ \hline	
		Non Specific Nouns & 9.2052 (5.881447, 9.13725) & 3.8627 (3.275567, 10.474625) & -0.37956 (-4.037218, 2.770237) \\ \hline
		LI Verbs & 4.0434 (10.84689, 31.53291) & 2.804 (10.91107, 67.41231) & 1.2836 (-11.06486, 49.00844)\\ \hline
	\end{tabular}
	\caption{\label{tab:table-name}RR T-tests vs GWB and DJT, T-statistic and 95\%confidence intervals}
	\end{center} 
\end{table}

\subsection{Longitudinal Analysis}
\subsubsection{Comparison of Ronald Reagan and George H.W. Bush}
We then looked at the data from a longitudinal perspective as we were interested in seeing whether we can track various language features and their progress over time. We ran a number of Pearsons correlations with number of days as a time reference (we calculated this as the total number of days from the first sample) and the dependent variables. We also calculated these Pearsons correlations with transcript index number as a time reference and the dependant variables as a replication of the Berisha and LIss study \cite{Berisha2015} and these results are described in the supplementary material.

For our controls, we found them to be stable for the most part with the main highlights being a decrease in Adverb usage for DJT (R=-0.36, p=0.049) and a steady but not severe decline in a number of variables for GWHB, namely total word count, unique words, low imageability words and verb usage.
\par 
For RR, his decline is more marked and more widespread through his language use. We noticed an significant increase in adverb (R=0.41, p=0.004) and pronoun usage (R=0.65, p\textless0.001), as well as a slight usage increase in Non-specific nouns(R=0.30, p=0.03). There was also a significant decrease in number of unique words (R=-0.56, p\textless0.001) and noun usage (R=-0.70, p\textless0.001). Also there was a significant decrease in adjective usage (R=-0.40, p=0.005) and a significant decrease in total word count (R=-0.31, p=0.03). 
\par
Given the number of hypotheses being tested, we felt it was necessary to control for false discovery rate and therefore Benjamii-Yekutieli procedure was applied to the results of the analysis. By applying this, we produced a final list of important features for each president. For RR, this was 22 features and for GHWB this was 14 features.  The features were generated from multiple sources and therefore there was some significant overlap between the features but for completeness we have included all features in Table 6.

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | l | c | c |}
		\hline
		& RR R-Squared  & GWB R-Squared \\ \hline
		ppron & \textbf{0.700***} & 0.007 \\ \hline
		social & \textbf{0.698***} & 0.204 \\ \hline
		NounsNormalised & \textbf{-0.689***} & -0.023 \\ \hline
		function & \textbf{0.670***} & -0.169 \\ \hline
		conj &\textbf{0.644***} & -0.452 \\ \hline
		PronounsNormalised & \textbf{0.631***} & 0.131 \\ \hline
		Analytic & \textbf{-0.626***} & -0.013 \\ \hline
		NN & \textbf{-0.601***} & -0.197 \\ \hline
		male & \textbf{0.585***} & 0.024 \\ \hline
		UniqueWords & \textbf{-0.578***} & -0.257 \\ \hline
		WDT & \textbf{-0.577***} & -0.173 \\ \hline
		shehe & \textbf{0.529*} & -0.005 \\ \hline
		VBZ & \textbf{-0.521*} & -0.157 \\ \hline
		JJ & \textbf{-0.518*} & -0.206 \\ \hline
		Fillers & -0.076 & \textbf{-0.359**} \\ \hline	
		EX & -0.193* & \textbf{-0.337**} \\ \hline	
		achieve & -0.248 & \textbf{0.334**} \\ \hline		
		\multicolumn{3}{@{}p{1.5in}}{\footnotesize * denotes p\textless 0.05} \\
    	\multicolumn{3}{@{}p{1.5in}}{\footnotesize ** denotes p\textless 0.01} \\
    	\multicolumn{3}{@{}p{1.5in}}{\footnotesize *** denotes p\textless 0.001} \\
	\end{tabular}
	\caption{\label{tab:table-name}Pearson Correlations for Features}
	\end{center} 
\end{table}

\subsubsection{Comparison of Ronald Reagan and Donald J. Trump}
\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | l | c | c |}
		\hline
		& RR R-Squared & DJT R-Squared \\ \hline
		ppron & \textbf{0.700***} & 0.007 \\ \hline
		social & \textbf{0.698***} & 0.204 \\ \hline
		NounsNormalised & \textbf{-0.689***} & -0.023 \\ \hline
		function & \textbf{0.670***} & -0.169 \\ \hline
		conj &\textbf{0.644***} & -0.452 \\ \hline
		PronounsNormalised & \textbf{0.631***} & 0.131 \\ \hline
		Analytic & \textbf{-0.626***} & -0.013 \\ \hline
		NN & \textbf{-0.601***} & -0.197 \\ \hline
		male & \textbf{0.585***} & 0.024 \\ \hline
		UniqueWords & \textbf{-0.578***} & -0.257 \\ \hline
		WDT & \textbf{-0.577***} & -0.173 \\ \hline
		shehe & \textbf{0.529*} & -0.005 \\ \hline
		VBZ & \textbf{-0.521*} & -0.157 \\ \hline
		JJ & \textbf{-0.518*} & -0.206 \\ \hline
		Fillers & -0.076 & \textbf{-0.359**} \\ \hline	
		EX & -0.193* & \textbf{-0.337**} \\ \hline	
		achieve & -0.248 & \textbf{0.334**} \\ \hline		
		\multicolumn{3}{@{}p{1.5in}}{\footnotesize * denotes p\textless 0.05} \\
    	\multicolumn{3}{@{}p{1.5in}}{\footnotesize ** denotes p\textless 0.01} \\
    	\multicolumn{3}{@{}p{1.5in}}{\footnotesize *** denotes p\textless 0.001} \\
	\end{tabular}
	\caption{\label{tab:table-name}Pearson Correlations for Features}
	\end{center} 
\end{table}
% Needs explanation

\subsection{Comparing linear models and non-linear models}
In terms of comparing the general additive model and linear models. We calculated the predicted residual error sum of squares (PRESS) statistic for each feature is a form of cross-validation used in regression analysis to provide a summary measure of the fit of a model to a sample of observations that were not themselves used to estimate the model. There were no significant differences (paired t-test: t=1.8875, df  = 21, p-value = 0.07299).

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | l | c | c |}
		\hline
		Feature & GAM Press Statistic & LM Press Statistic \\ \hline
		ppron & 18.591104613 & 20.001055542 \\ \hline
		social & 21.681732112 & 24.906646115  \\ \hline
		NounsNormalised & 0.002818941 & 0.002797798\\ \hline
		function & 46.546071705 & 46.031993602 \\ \hline
		conj & 8.812330188 & 8.354974826 \\ \hline
		PronounsNormalised & 0.002284437 & 0.002356129\\ \hline
		Analytic & 2272.200519272 & 2239.224863610 \\ \hline
		Pronoun & 44.557282749 & 46.894576916 \\ \hline		
		NN & 101685.870974277 & 96485.575598505 \\ \hline
		male & 5.191815443 & 5.278559152 \\ \hline
		UniqueWords & 255114.636660673 & 238042.427437454 \\ \hline
		WDT & 1987.149134694 & 1958.781236655 \\ \hline
		nouns & 276723.222762842 & 262192.120091097 \\ \hline
		nouns/100 & 27.670762078 & 26.219212009 \\ \hline
		UniqueStems & 170450.645283172 & 166302.725530118 \\ \hline
		shehe & 5.188136314 & 5.265385185 \\ \hline
		VBZ & 13889.871831536 & 13721.175422577 \\ \hline
		JJ & 24950.440222526 & 24692.043386181 \\ \hline
		article & 13.500676675 & 13.468083961 \\ \hline	
		Adjectives & 30210.006270270 & 29865.240707991 \\ \hline	
		Adjectives.100 & 3.020898615 & 2.986524071 \\ \hline
		Dic & 40.099289973 & 38.501060715 \\ \hline	
	\end{tabular}
	\caption{\label{tab:table-name}Comparison of GAM and Linear Model using the PRESS statistic}
	\end{center} 
\end{table}

@@

@@

Our next hypothesis involved exploring whether linear models were the most appropriate way to track this longitudinal data. In our analysis, we found there was no significant difference in terms of mean sum of squares errors between the linear model and the generalised additive model. However this does not mean that these are equivalent. For example, consider Figure 2. We compare a model fit to a linear model with a generalised additive model which aims to model the data more closely. It is clear to see that both models can track the movements of language use over time however, the linear model is quite a rigid model which does not allow for different inflection points. We can see that RR's decline is not linear, and that there are periods of relative stability and some periods where the decline is more severe. We argue that these more nuanced changes over time can only be modelled by a model such as a GAM.

@@

\section{Conclusions}\label{conclusions}
The results of this work show that we can track a person's use of language through time in a number of ways and that it is possible for an individual to be his or her own control. This is important as it means the heterogenous nature of the MCI population does not impact results as much as if we were comparing those with group of MCI patients with a group of controls. Equally, it would be helpful to have controls to ascertain what would be usual to expect in the decline of language in a healthy older adult. 
\par 
The results of this work also identify some clues as to what could work in a clinical setting, or how we might be able to collect data in a way that accurately tracks language decline without making incorrect assumptions. 
\par
From a clinical perspective, we can see that using samples (albeit for this dataset) is able to track languages changes over a given time frame. This potentially means that we can use a similar methodology to collect regular language samples in settings such as memory clinics and GP's surgeries and even potentially in the people's homes and that these language samples, have the potential to act as an early warning sign for Mild Cognitive Impairment that potentially will identify patients at risk of developing Alzheimer's Disease.
\bibliographystyle{ieeetr}
\bibliography{main}

>>>>

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=6cm, height=5cm]{plots/comparison1.png}
	\caption{Comparing linear model with a non-linear model}
\end{figure}

@@

\subsection{Longitudinal Analysis}
As with Berisha and Liss\cite{Berisha2015}, we analysed the transcripts over time in order to determine if there was an underlying temporal trend that may indicate that different parts of language increase or decrease over time. In AD, cognitive abilities are said to deteriorate over time and therefore it may be useful to analyse temporal trends that indicate a potential deterioration of cognition. In order to get a accurate measure of deterioration over time, we time stamped each transcript in relation to number of days from the first transcript to the date of the transcript being analysed as the press conferences were not evenly spaced out during the presidencies. We then applied a pearsons product moment correlation to examine whether these language features increased or decreased significantly over time. Finally, for the features we found significant for RR we compared these results with GHWB and DJT using ??? to look at whether the difference was significant given the sample size. 
\par
We looked also compared the results of linear models and a non linear model in terms of best fit. The non linear model we used was the generalised additive model \cite{Hastie1986}.
There is evidence to support the idea that language deterioration. In figure 1, we look at the difference of between a linear model of Reagan's use of personal pronouns vs a non linear model. Whilst they are broadly similar in that they track an increase in his use of personal pronouns, they do not capture the variation in this increase as closely as a non linear model might. We therefore decided to apply this to all the features we extracted.

@@

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{3cm} | p{3cm} | p{3cm} | p{3cm} |}
		\hline
		& RR v GHWB & RR v DJT & GHWB v DJT \\ \hline
		Total Words & 6.6751 (479.77, 1134.64) & 5.352 (766.95, 2383.60) & 2.4774 (-74.34, 1592.44) \\ \hline
		Unique Stems & 10.878 (149.73, 244.04) & 10.11 (251.17, 437.85) & 4.148 (51.53, 244.26) \\ \hline
		Mean Length of Utterance & 16.175 (3.73, 5.19) & 25.084 (8.21, 10.17)  & 13.484 (3.78, 5.66) \\ \hline	
		Non Specific Nouns & 9.2052 (5.35, 9.66) & 3.8627 (2.06, 11.69) & -0.37956 (-5.21, 3.95) \\ \hline
		LI Verbs & 4.0434 (7.54, 34.84) & 2.804 (0.78, 79.55) & 1.2836 (-21.35, 59.30)\\ \hline
	\end{tabular}
	\caption{\label{tab:table-name}RR T-tests vs GWB and DJT, T-statistic and 99\% confidence intervals rounded to 2.d.p}
	\end{center} 
\end{table}

\subsection{Longitudinal Analysis}
\subsubsection{Comparison of Ronald Reagan and George H.W. Bush}
We then looked at the data from a longitudinal perspective as we were interested in seeing whether we can track various language features and their progress over time. We ran a number of Pearsons correlations with number of days as a time reference (we calculated this as the total number of days from the first sample) and the dependent variables. We also calculated these Pearsons correlations with transcript index number as a time reference and the dependant variables as a replication of the Berisha and LIss study \cite{Berisha2015} and these results are described in the supplementary material.
\par 
For our controls, we found them to be stable for the most part with the main highlights being a decrease in Adverb usage for DJT (R=-0.36, p=0.049) and a steady but not severe decline in a number of variables for GWHB, namely total word count, unique words, low imageability words and verb usage.
\par 
For RR, his decline is more marked and more widespread through his language use. We noticed an significant increase in adverb (R=0.41, p=0.004) and pronoun usage (R=0.65, p\textless0.001), as well as a slight usage increase in Non-specific nouns(R=0.30, p=0.03). There was also a significant decrease in number of unique words (R=-0.56, p\textless0.001) and noun usage (R=-0.70, p\textless0.001). Also there was a significant decrease in adjective usage (R=-0.40, p=0.005) and a significant decrease in total word count (R=-0.31, p=0.03). 
\par
Given the number of features being compared, we felt it was necessary to control for false discovery rate and therefore Benjamii-Yekutieli procedure was applied to the results of the analysis. By applying this, we produced a final list of important features for each president. For RR, this was 22 features and for GHWB this was 14 features.  The features were generated from multiple sources and therefore there was some significant overlap between the features. Where this was the case, for example total nouns and nouns normalised,  we selected selected the feature that best controlled for document length (see Table 6). We also removed any features which calculated the use of punctuation as this was added later by the transcriptionist. This reduced the number of features that were significant for RR to 14 and for GHWB to 3. The majority of these features was significantly different in terms of movement over time when comparing RR to GHWB. The three features that were significant for GHWB in terms of correlation coefficient were not significant when comparing them with RR.

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | l | c | c |}
		\hline
		& RR R-Squared  & GWB R-Squared \\ \hline
		ppron & \textbf{0.700*} & 0.007 \\ \hline
		social & \textbf{0.698*} & 0.204 \\ \hline
		NounsNormalised & \textbf{-0.689*} & -0.023 \\ \hline
		function & \textbf{0.670*} & -0.169 \\ \hline
		conj &\textbf{0.644*} & -0.452 \\ \hline
		PronounsNormalised & \textbf{0.631*} & 0.131 \\ \hline
		Analytic & \textbf{-0.626*} & -0.013 \\ \hline
		NN & \textbf{-0.601*} & -0.197 \\ \hline
		male & \textbf{0.585*} & 0.024 \\ \hline
		UniqueWords & \textbf{-0.578*} & -0.257 \\ \hline
		WDT & \textbf{-0.577*} & -0.173 \\ \hline
		shehe & \textbf{0.529*} & -0.005 \\ \hline
		VBZ & \textbf{-0.521*} & -0.157 \\ \hline
		JJ & \textbf{-0.518*} & -0.206 \\ \hline
		Fillers & -0.076 & \textbf{-0.359*} \\ \hline	
		EX & -0.193* & \textbf{-0.337*} \\ \hline	
		achieve & -0.248 & \textbf{0.334*} \\ \hline		
		\multicolumn{3}{@{}p{1.5in}}{\footnotesize * denotes p\textless 0.05} \\
	\end{tabular}
	\caption{\label{tab:table-name}Pearson Correlations for Features for RR and GHWB}
	\end{center} 
\end{table}

\subsubsection{Comparison of Ronald Reagan and Donald J. Trump}
We then used DJT as another control to see our results from our comparison with GHWB held true. We found that the same 53 features were significantly different between RR and DJT and all of the features which we deemed significant in terms of increase or decrease were included. This confirms our results from our original comparison.

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | l | c | c |}
		\hline
		& RR R-Squared & DJT R-Squared \\ \hline
		ppron & \textbf{0.700*} & -0.196 \\ \hline
		social & \textbf{0.698*} & 0.032 \\ \hline
		NounsNormalised & \textbf{-0.689*} & 0.194 \\ \hline
		function & \textbf{0.670*} & -0.244 \\ \hline
		conj &\textbf{0.644*} & -0.187 \\ \hline
		PronounsNormalised & \textbf{0.631*} & 0.002 \\ \hline
		Analytic & \textbf{-0.626*} & 0.367 \\ \hline
		NN & \textbf{-0.601*} & 0.116 \\ \hline
		male & \textbf{0.585*} & 0.043 \\ \hline
		UniqueWords & \textbf{-0.578*} & 0.204 \\ \hline
		WDT & \textbf{-0.577*} & -0.106 \\ \hline
		shehe & \textbf{0.529*} & 0.065 \\ \hline
		VBZ & \textbf{-0.521*} & 0.136 \\ \hline
		JJ & \textbf{-0.518*} & 0.063 \\ \hline	
		\multicolumn{3}{@{}p{1.5in}}{\footnotesize * denotes p\textless 0.05} \\
	\end{tabular}
	\caption{\label{tab:table-name}Pearson Correlations for Features for RR and DJT}
	\end{center} 
\end{table}
% Needs explanation

\subsection{Comparing linear models and non-linear models}
In terms of comparing the general additive model and linear models. We calculated the predicted residual error sum of squares (PRESS) statistic for each feature. The PRESS statistic
is a form of cross-validation used in regression analysis to provide a summary measure of the fit of a model to a sample of observations that were not themselves used to estimate the model. There were no significant differences when comparing the models (paired t-test, t = 1.8875, df = 21, p = 0.07299).

\begin{table}[H]
\centering
\begin{tabular}{lrr}
  \hline
feature & gamPRESS & lmPRESS \\ 
  \hline
ppron & 18.59 & 20.00 \\ 
  social & 21.68 & 24.91 \\ 
  NounsNormalised & 0.00 & 0.00 \\ 
  function. & 46.55 & 46.03 \\ 
  conj & 8.81 & 8.35 \\ 
  PronounsNormalised & 0.00 & 0.00 \\ 
  Analytic & 2272.20 & 2239.22 \\ 
  NN & 101685.87 & 96485.58 \\ 
  male & 5.19 & 5.28 \\ 
  UniqueWords & 255114.64 & 238042.43 \\ 
  WDT & 1987.15 & 1958.78 \\ 
  Nouns/100 & 27.67 & 26.22 \\ 
  UniqueStems & 170450.65 & 166302.73 \\ 
  shehe & 5.19 & 5.27 \\ 
  VBZ & 13889.87 & 13721.18 \\ 
  JJ & 24950.44 & 24692.04 \\ 
  article & 13.50 & 13.47 \\ 
  Adjectives.100 & 3.02 & 2.99 \\ 
  Dic & 40.10 & 38.50 \\ 
   \hline
\end{tabular}
\end{table}

@@

Another we did was to look at the difference in decline between the presidents. We compared both RR v GHWB and RR v DJT. We found that for those features that were significant in terms of language decline for RR, they were significantly different for both our controls. This does appear to mean that for RR, his language changed in a considerably different way as he got older and we could attribute this decline to a pre-clinical stage of alzheimer's disease.

@@

Our next hypothesis involved exploring whether linear models were the most appropriate way to track this longitudinal data. In our analysis, we found there was no significant difference in terms of mean sum of squares errors between the linear model and the generalised additive model. However this does not mean that these are equivalent. For example, consider Figure 2. We compare a model fit to a linear model with a generalised additive model which aims to model the data more closely. It is clear to see that both models can track the movements of language use over time however, the linear model is quite a rigid model which does not allow for different inflection points. We can see that RR's decline is not linear, and that there are periods of relative stability and some periods where the decline is more severe. We argue that these more nuanced changes over time can only be modelled by a model such as a GAM. In Le's et al's paper, he looks at the written work of Iris Murdoch who would go on to receive a diganosis of Alzheimer's disease. He describes something called 'Murdoch's trough' which is a period of decline in her late 40's and early 50's, however there was a significant period following this of an improvement in her linguistic ability before a subsequent severe decline in her last novel. Figure 2 illustrates this, as well as what a non-linear model would look like vs a linear model.
\begin{figure}[H]
	\centering
	\includegraphics[width=6cm, height=5cm]{plots/IMStems.png}
	\caption{Measuring unique word stems in the novels of Iris Murdoch over time}
\end{figure}
We can clearly see that a linear model is insufficient for modelling this particular case and whilst Iris Murdoch may be an atypical case, we can see the value in using non-linear models that may be able to map these unusual cases whilst retaining the ability to illustrate fluctuations in language use.

@@

\section{Conclusions}\label{conclusions}
The results of this work show that we can track a person's use of language through time in a number of ways and that it is possible for an individual to be his or her own control. This is important as it means the heterogenous nature of the MCI population does not impact results as much as if we were comparing those with group of MCI patients with a group of controls. Equally, it would be helpful to have controls to ascertain what would be usual to expect in the decline of language in a healthy older adult. We were able to identify a number of linguistic features that merit further exploration in those with MCI or AD. 
\par 
The results of this work also identify some clues as to what could work in a clinical setting, or how we might be able to collect data in a way that accurately tracks language decline without making incorrect assumptions. 
\par
From a clinical perspective, we can see that using samples (albeit for this dataset) is able to track languages changes over a given time frame. This potentially means that we can use a similar methodology to collect regular language samples in settings such as memory clinics and GP's surgeries and even potentially in the people's homes and that these language samples, have the potential to act as an early warning sign for Mild Cognitive Impairment that potentially will identify patients at risk of developing Alzheimer's Disease.
\bibliographystyle{ieeetr}
\bibliography{main}
